Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-08-27 main.Rmd
Clean up of Rmd file

authors:
- name: Sayani Gupta
  affiliation: Department of Econometrics and Business Statistics, Monash University
  thanks: "Email: Sayani.Gupta@monash.edu"

- name: Rob J Hyndman
  affiliation: Department of Econometrics and Business Statistics, Monash University

- name: Dianne Cook
  affiliation: Department of Econometrics and Business Statistics, Monash University

- name: Antony Unwin
  affiliation: University of Augsburg

@@

abstract: |
 Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes several classes of time deconstructions using linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months with respect to a baseline. Cyclic time granularities can be circular such as hour of the day, quasi-circular such as day of the month, and aperiodic such as public holidays. The hierarchical structure of granularities creates a nested ordering. Hour of the day and second of the minute are single-order-up. Hour of the week is multiple-order-up, because it passes over day of the week. Methods are provided for creating all possible granularities for a time index.  A recommendation algorithm provides an indication whether a pair of granularities can be meaningfully examined together, called a harmony or when they cannot, called a clash.

 The time granularities can be used to create visualizations of the data to explore for periodicities, associations and anomalies. The granularities can be considered to be categorical variables (ordered or unordered) which induces a grouping of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables. A recommendation of appropriate distribution display is provided.

 The methods are implemented in the open source R package `gravitas`. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow (@Grolemund2018-po), and the probability distributions can be examined using the range of graphics available in `ggplot2` [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes,amsthm}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, comment = "#>",
  fig.path = "figure/", fig.align = "center", fig.show = "hold",
  cache = TRUE, cache.path = "cache/",
  out.width = ifelse(is_html_output(), "100%", "\\textwidth")
)
# knitr::opts_knit$set(root.dir = here::here())
```

@@

```{r load}

```

@@

<!--temporal granularities and why should we care -->

<!-- problem of analyzing data with such finer scales and classical ways to deal with it -->

@@

<!-- # existing ways and challenges -->

@@

<!--motivation -->

@@

<!-- at finer or coarser scales can be benefited from the approach of Exploratory Data Analysis (EDA). EDA calls for utilizing visualization and transformation to explore data systematically. It is a process of generating hypothesis, testing them and consequently refining them through investigations. -->

<!--existing closely related work and how ours relate -->
 <!-- This paper utilizes the nestedness of time granularities to obtain multiple-order-up granularities from single-order-up ones. -->

<!--actions on multiple order up - usage -->

<!-- Finally, visualizing data across single/multiple-order-up granularities help us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. However, this approach can lead to a myriad of choices all of which are not useful. Analysts are expected to alliteratively visualize these choices for exploring possible patterns in the data. But too many choices might leave him bewildered.  -->

<!-- challenges and how you are dealing wih it -->
This work provides tools for systematically exploring bivariate granularities within the tidy workflow. In particular, we

@@

<!-- Pairs of granularities can be categorized as either a *harmony* or *clash*, where = are pairs of granularities that aid exploratory data analysis, and clashes are pairs that are incompatible with each other for exploratory analysis. -->

@@

<!-- general calendar categorization -->
<!-- Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Classes of time abstractions may be based on arrangement or hierarchical order of time, discussed in further details in \autoref{sec:arrangement} and \autoref{sec:order}. -->

@@

<!-- Arrangement: Linear vs. Circular vs. aperiodic {#sec:arrangement} -->

<!-- The arrangement of the time domain can result in deconstructing time in linear, circular or aperiodic (nearly circular) ways.  -->

<!-- There have been several attempts to provide a framework for formally characterizing time granularities. One of the first attempts was provided by @Bettini1998-ed who proposed following definitions:-->

<!-- <!-- Representation with tsibble -->
<!-- Suppose we have a tsibble with a time index in one column and keys and variables in other columns. A time domain, as defined by [@Bettini1998-ed], is essentially a mapping of row numbers (the index set) to the time index. A linear granularity is a mapping of row numbers to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. -->

<!-- <!-- What is calendar categorization -->
<!-- What we need to add to this are additional categorizations of time that are not linear granularities and are useful to represent periodicity. Examples include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc. Many of these are circular, such as day-of-week, time-of-day. Some are nearly circular such as day-of-month. Some are irregular such as working day/non-working day. Let's call all of these "calendar categorizations". Anything that maps a time index to a categorical variable can be considered a **calendar categorization**.  -->

<!-- What is circular granularity and some examples -->
<!-- We specify the circular categorizations using modular arithmetic and call these **circular granularities**. The number of categories is essentially the periodicity of a circular time granularity. For example, suppose the time index is in minutes, and let $n_i$ be the number of categories created by the circular granularity $C_i$. Then the following categorizations can be computed. -->

<!-- In the following sections, we provide a formalism to all of these abstractions and identify their relationships and symbolic representations.  -->

<!-- ### Linear {#sec:linear-gran-def} -->

\newtheorem{definition}{Definition}

\begin{definition}\label{def:definition}
A {\bf time domain} is a pair $(T; \le)$ where $T$ is a non-empty set of time instants and $\le$ is a total order on $T$.
\end{definition}

The time domain is assumed to be *discrete*, and there is unique predecessor and successor for every element in the time domain except for the first and last.

\begin{definition}\label{def:index set}
There is a unique {\bf index set}, $Z=\{z: z \in \mathbb{Z}_{\geq 0}\}$, which map the time instants to a set of non-negative integers. Often, this is thought of as $t=0, \dots, T$.
\end{definition}

\begin{definition}\label{def:linear}
A {\bf linear granularity} is a mapping $G$ from the integers (the index set, $Z$) to subsets of the time domain such that:
  (1) if $i < j$ and $G(i)$ and $G(j)$ are non-empty, then each element of $G(i)$ is less
than all elements of $G(j)$; and
  (2) if $i < k < j$ and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty.
Each non-empty subset $G(i)$ is called a {\bf granule}.
\end{definition}

\noindent This implies that the granules in a linear granularity are non-overlapping, continuous and ordering is maintained. The indexing for each granule can also be associated with textual representation, called the label.  A discrete time model often uses a fixed smallest linear granularity named by @Bettini1998-ed **bottom granularity**.<!--The label can be mapped to index for input and the index can later be mapped to this label for output. When the label set is exactly the integers, the granularities are called "full-integer" labelled granularity, such as in Definition \ref{def:linear}. Some calendar operations will only work on full-integer labelled granularities, while others can be implemented using more flexible labelings.--> \autoref{fig:linear-time} illustrates the linear time granularities. Here, "hour" is the bottom granularity and "day", "week", "month" and "year" are linear granularities which maps from index set to subsets of the hourly time domain. <!--For example, day($0$) =  {hour($0$), hour($1$), hour($2$), \dots, hour($23$)}, day($1$) =  {hour($24$), hour($25$), hour($26$), \dots, hour($47$)} and so on.--> If we have "hour" running from $\{0, 1, \dots,t\}$, we will have "day" running from $\{0, 1, \dots, \lfloor t/24\rfloor\}$. These linear granularities are ordered with ordering guided by the index set which is a set of integers. Hence, they are uni-directional and non-repeating.

 <!--, each of which are non-repeating in nature and arranged from past to future. If we consider $T$ hours, we will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on.-->

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="(ref:linear-time)"}
```

@@

<!--@Bettini1998-ed discusses the relationships between linear time granularities, periodicities in time, and the structure of a calendar.-->

@@

\begin{definition}\label{def:finerthan}
A linear granularity $G$ is {\bf finer than} a linear granularity $H$, denoted $G \preceq H$, if for each index $i$, there exists an index $j$ such that
$G(i) \subset H(j).$
\end{definition}

\begin{definition}\label{def:groupsinto}
A linear granularity $G$ {\bf groups into} a linear granularity $H$, denoted
$G \trianglelefteq H$, if for each index $j$ there exists a (possibly infinite) subset $S$ of the integers such that $H(j) = \bigcup_{i \in S}G(i).$
\end{definition}

<!-- CONTENT CUT -->
<!-- These definitions are illustrated in \autoref{fig:groups-finer}, where both day $\trianglelefteq$ week, and day $\preceq$ week, hold true. However, weekends are finer than weeks but do not group into weeks. And weekends neither groups into or is finer than public holidays. Consider another example where day $\trianglelefteq$ month. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions. -->
\newtheorem*{example}{Example}

\begin{example}
{\rm Both $day \trianglelefteq week$ and $day \preceq week$ holds, since every granule of $week$ is the union of some set of granules of day and each day is a subset of a $week$. Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year, if leap years are ignored. This grouping period becomes 4 and 400 years with the inclusion of leap years and leap seconds respectively.}
\end{example}

<!-- CONTENT CUT -->
<!-- # ```{r groups-finer, echo=FALSE, out.width = "100%", fig.cap="Illustrations of groups into and finer than relationships. Weekends are finer than weeks but do not group into weeks. And weekends neither groups into or finer than public holidays"} -->
<!-- knitr::include_graphics("Figs/groupsandfiner.png") -->
<!-- ``` -->

\begin{definition}\label{def:periodical}
A granularity $G$ is {\bf periodical} with respect to a granularity $H$ if:
(1) $G \trianglelefteq H$; and
(2) there exist $R$, $P \in \mathbb{Z}_+$, where $R$ is less than the number of granules of $H$, such that for all $i \in \mathbb{Z}$, if $H(i) = \bigcup_{j \in S}G(j)$ and $H (i + R) \neq \phi$ then $H (i + R) = \bigcup_{j \in S} G(j + P)$.
\end{definition}

<!-- CONTENT CUT -->
<!-- A granularity $H$ which is periodical with respect to $G$ is specified by: -->
<!-- (i) the $R$ sets of indexes of $G$, ${S_0,\dots,S_{R-1}}$ describing the granules of $H$ within one period; -->
<!-- (ii) the value of $P$; -->
<!-- (iii) the indexes of first and last granules in $H$, if their value is not infinite. -->

@@

<!-- CONTENT CUT -->
<!-- Intuitively, the definition requires that all granules of $H$ within the span of time between two exceptions have the same periodical behavior, characterized by $R$ and $P$. -->

<!-- # ```{r group-periodical, echo=FALSE, out.width = "100%", fig.cap="Linear granularity $B$ groups periodically into linear granularity $G$ with $P = 5$ and $R = 2$, alternating between combining 2 and 3 granules of $B$."} -->
<!-- # knitr::include_graphics("Figs/groups_periodically.png") -->
<!-- # ``` -->

<!-- CONTENT CUT -->
<!-- \begin{definition}[Groups quasi-periodically]\label{def:quasi} -->
<!-- A granularity $H$ is quasi-periodical with respect to a granularity $G$ if: -->
<!-- (1) $G \trianglelefteq H$; and -->
<!-- (2) there exist a set of intervals $E_1,\dots,E_z$ (the granularity exceptions) and positive integers $R$, $P$, where $R$ is less than the minimum of the number of granules of $H$ between any two exceptions, such that for all $i \in \mathbb{Z}$, if $H(i) = \bigcup_{k \in [0,k]} G(j_r)$  and $H (i + R) \neq \phi$ and $i + R < min(E)$, where $E$ is the closest existing exception after $H (i)^2$, then $H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P)$. -->
<!-- \end{definition} -->

<!-- # ```{r quasi-group-periodical, echo=FALSE, out.width = "100%", fig.cap="Linear granularity $B$ groups quasi-periodically into linear granularity $G$ with $P = 5$ and $R = 2$, alternating between combining 2 and 3 granules of $B$ with $E1$ as the exception."} -->
<!-- # knitr::include_graphics("Figs/quasi_groups_periodically.png") -->
<!-- # ``` -->

<!-- **Example**:  Example showing periodic and quasi-periodic linear granularities are displayed in \autoref{fig:group-periodical} and \autoref{fig:quasi-group-periodical} respectively. Each granule is represented by a box. In these figures, the granules that overlap share elements from the underlying time domain. In both of these cases, each granule of $G$ is made up of two and three granules of $B$ alternately. So the period is $P = 5$ and number of granules of $G$ in each period is given by $R=2$. The only difference in \autoref{fig:quasi-group-periodical} is that there is an exception after the first granule of $G$ which consists of one granule of $B$. Except for this exception $E1$, $B$ groups periodically into $G$.  -->

\begin{example}
{\rm In a Gregorian calendar without leap years we could say $day$ groups periodically into month with the period $P = 365$ and the number of granules of $month$ in each period given by $R=12$. In a Gregorian calendar with leap years, $day$ groups quasi-periodically into $month$ with the exceptions of the time domain corresponding to $29^{\text{th}}$ February of any year.}
\end{example}

<!-- CONTENT CUT -->
<!-- \begin{definition}[Bottom granularity]\label{def:B} -->
<!-- Given a granularity order relationship g-rel and a set of granularities having the same time domain, a granularity B in the set is a bottom granularity with respect to g-rel, if B g-rel H for each granularity H in the set. -->
<!-- \end{definition} -->

<!-- Here, g-rel could be any relationship defined between granularities as defined in @Bettini1998-ed. Given the set of all granularities defined over the time domain $(\mathbb{Z}; <)$, and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$. -->

\begin{definition}\label{def:order}
The {\bf order} of a linear granularity is the level of coarseness associated with a linear granularity. A linear granularity G will have lower order than H if each granule of G is composed of lower number of granules of bottom granularity than each granule of H.
\end{definition}

\begin{example}
{\rm For two linear granularities $G$ and $H$, if $G$ {\em groups into} or {\em finer than} $H$ then $G$ is of lower order than $H$. Moreover, if we consider bottom granularity as day, linear granularity week will have lower order than month since each week consist of less number of days than each month. Here neither  week {\em groups into} month or  week {\em finer than} month, but their relative order could be determined.}
\end{example}

<!-- that the aperiodic linear granularities cyclic granularity $C_{G(l),G(m)}$ is either circular or quasi-circular.  -->

<!--so that the the underlying time domain can be described as a sequence of non-decomposable, consecutive time intervals of some fixed, minimal duration called _chronons_ (@Jensen1998-qn). Chronons may be grouped into granules of a bottom granularity and -->

@@

<!-- CONTENT CUT -->
<!-- For a coarser granularity $G_2$ and a finer one $G_1$, a _grouping operation_ partitions some granules of $G_1$  to form a granule of $G_2$. For example, $\text{minute} = \text{Group}_{60}(\text{second})$; $\text{hour} = \text{Group}_{60}(\text{minute})$. The _altering-tick operation_ generates new full-integer labelled granularity by periodically expanding or shrinking granules of $G_2$ in terms of granules of $G_1$. For example, <!--due to the uneven length of months, we need to use the altering-tick operation to compute months from days. We can consider a granularity "pseudomonth" which computes months only for non-leap years by grouping 31 days, and then shrink April, June, September and November by 1 day, and shrink February by 3 days [@Ning2002-tf]. For more variations of calendar algebra operations, see @Ning2002-tf.

<!-- CONTENT CUT -->
<!-- The calendar algebra consists of two kinds of operations: grouping-oriented and granule-oriented operations. The grouping-oriented operations combine certain granules of a granularity together to form the granules of the new granularity. For example, consider a calendar with only two linear granularities: "minute" and "hour", where "hour" is generated by grouping every $60$ "minutes". -->
<!-- The granule-oriented operations do not change the granules of a granularity, but rather select which granules should remain in the new granularity. For example, one can define the granularity "Monday" and hence select only Mondays from linear granularity "day". -->

<!-- \begin{definition}[Labelled granularity]\label{def:labelgran} -->

<!-- CONTENT CUT -->
<!-- The notion of a labelled granularity is introduced in @Bettini2000-qk to simplify the calendar algebra. A labelled granularity is a pair $(L, G)$, where the label set $L$ is a subset of the integers instead of the set of integers and -->
<!-- follows the conditions (1) and (2) in \ref{def:linear} where $i$, $j$, and $k$ are in $L$. When the label set $L$ is exactly the integers, the granularities are called "full-integer" labelled granularity, such as in Definition \ref{def:linear}.<!-- However, the set $L$ of labels can be an arbitrary subset of (possibly non contiguous) integers and could be used to identify granules.-->

<!-- $G$ is a mapping from $L$ to the subsets of the time domain such that: -->

<!--   1. if $i<j$ , where $i$ and $j$ are in $L$, and $G(i)$ and $G(j)$ are non-empty, then each element in $G(i)$ is less than all the elements of $G(j)$; and -->
<!--   2. if $i<k<j$ , where $i$, $j$, and $k$ are in $L$, and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty. -->

\begin{definition}\label{def:grouping_operation}
Let $G_1$ be a full-integer labelled granularity, and $m$ a positive integer. {\bf The grouping operation} Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labelled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$
The grouping operation Group$_m(G)$, provides a new granularity whose granules are the integer part of the quotient when the integer set spanned by linear granularity $G$ is divided by $m$, such that, $G_2(i) = \lfloor G_1(j)/m \rfloor$.
\end{definition}

@@

<!-- Alternatively, given any integer $G_1(j)$ and a positive integer $m$, if there exists unique integers $G_2(i)$ and $R$ such that $G1(j) = m * G2(i) + R$ where $0 ≤ R < m$, then $G_2$ is a granularity that is formed -->

<!-- CONTENT CUT altering-tick -->
<!-- \begin{definition}\label{altering-tick} -->
<!-- Let $G_1$, $G_2$ be full-integer labelled granularities, and $l$, $k$, $m$ integers, with $G_1 \trianglelefteq G_2$, $G_1\preceq G_2$, and $1 \leq l \leq m$. {\bf The altering-tick operation} $\text{Alter}^{m}_{l,k} (G_2, G_1)$ operation modifies the granules of $G_1$ so that the $l^{\text{th}}$ granule of each period has $k$ additional granules of $G_2$. -->
<!-- \end{definition} -->

<!-- \begin{example} -->
<!-- The altering-tick operation can be used to compute the granularity $pseudomonth$, for the uneven lengths of $month$, by grouping 31 granules of $day$, followed by shrinking April, June, September and November by 1 granule of $day$, and shrinking February by 3 granules of $day$ for non-leap years.  -->
<!-- \end{example} -->

@@

<!-- CONTENT CUT -->
<!-- \begin{align*} -->
<!-- \text{pseudomonth} = &\text{Alter}^{12}_{11,-1}(\text{day}, \text{Alter}^{12}_{9,-1}(\text{day}, \\ -->
<!--   &~~\text{Alter}^{12}_{6,-1}(\text{day}, \text{Alter}^{12}_{4,-1)}(\text{day}, \\ -->
<!--   &~~~\text{Alter}^{12}_{(2,-3)}(\text{day}, \text{Group}_{31}(\text{day})))))), -->
<!-- \end{align*} -->

@@

<!-- sub: regular and irregular mapping leads to circular and quasi-circular granularities -->

Cyclic granularities represent cyclical repetitions in time. They can be thought of as additional categorizations of time that are not linear.<!--Weeks is an example of a cyclical granularity because every 7 days there's another Monday.--> Cyclic granularities can be constructed by operating on two linear granularities. Cycles can be either _regular_, called **circular**, or _irregular_, **quasi-circular** when these two linear granularities relate periodically. <!-- with every granule of coarser linear granularity being made up of same (regular) or different (irregular) number of granules of finer linear granularity. For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31). A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Cyclic time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mappings are irregular. Examples of common circular granularities include hour of the day and day of the week, whereas examples for quasi-circular granularities are day of the month and week of the month.-->

<!--sub:How to relate linear granularities to tsibble and manipulation of indices required for circular and quasi-circular granularities -->

@@

\begin{definition}\label{def:circular}
A {\bf circular granularity} $C_{B, G}$ relates a linear granularity $G$ to the bottom granularity B, if
\begin{equation} \label{eq:circular-gran}
\begin{split}
C_{B, G}(z) & = z\mod P(B, G) \quad \forall z \in \mathbb{Z}_{\geq 0} \\
\end{split}
\end{equation}
where
z denotes the index set,
B denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G$ with regular mapping, and $P \equiv P(B, G)$ is the number of granules of $B$ in each granule of $G$, also called the period of the grouping $(B, G)$.
\end{definition}

<!-- The label can be mapped to index for input and the index can later be mapped to this label for output. When the label set is exactly the integers, the granularities are called "full-integer" labelled granularity, such as in Definition \ref{def:linear}. Some calendar operations will only work on full-integer labelled granularities, while others can be implemented using more flexible labelings. -->

\begin{example}
{\rm \autoref{fig:circular-dow} illustrates the linear and the corresponding cyclical granularities. Cyclical granularities can be considered to be cutting the linear granularity into pieces, and stacking them to match the cycles (as shown in b). $B, G, H$ (day, week, fortnight, respectively) are linear granularities. The circular granularity $C_{B , G}$ (day-of-week)is constructed from $B$ and  $G$. The circular granularity $C_{B, H}$ (day-of-fortnight) is constructed from $B$ and $H$. These overlapping cyclical granularities share elements from the linear granularity. Each of $C_{B , G}$ and $C_{B , H}$ consist of repeated patterns  $\{0, 1, 2, \dots, 6\}$ and $\{0, 1, 2, \dots, 13\}$ with $P=7$ and $P=14$ respectively. Each circular granularity can use descriptive label mappings. Suppose ${L}$ is a label mapping that defines an unique label for each index $l \in \{ 0,1,\dots, (P-1)\}$, then the label mapping $L$ for $C_{B, G}$ can be defined as
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sun}, \text{Mon}, \dots, \text{Sat}\}
$$
or
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sunday}, \text{Monday}, \dots, \text{Saturday}\}
$$
for example.}
\end{example}

<!-- # ```{r circular-gran, echo=FALSE,out.width = "100%",fig.cap="Relating circular granularities and bottom linear granularity "} -->
<!-- # knitr::include_graphics("Figs/circular-gran-diagram.png") -->
<!-- # ``` -->

```{r circular-dow,echo=FALSE,out.width = "100%",fig.cap= "(ref:circular-dow)"}

```

@@

<!-- All circular granularities can be expressed in terms of the bottom linear granularity with equation \autoref{eq:eq2}. -->

>>>>

authors:
- name: Sayani Gupta
  thanks: "Email: Sayani.Gupta@monash.edu"
  affiliation: Department of Econometrics and Business Statistics, Monash University, Australia

- name: Rob J Hyndman
  affiliation: Department of Econometrics and Business Statistics, Monash University, Australia

- name: Dianne Cook
  affiliation: Department of Econometrics and Business Statistics, Monash University, Australia

- name: Antony Unwin
  affiliation: University of Augsburg, Germany

@@

abstract: |
 Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes classes of time deconstructions using linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Cyclic time granularities can be circular such as hour-of-the-day, quasi-circular such as day-of-the-month, and aperiodic such as public holidays. The hierarchical structure of granularities creates a nested ordering: hour-of-the-day and second-of-the-minute are single-order-up. Hour-of-the-week is multiple-order-up, because it passes over day-of-the-week. Methods are provided for creating all possible granularities for a time index. A recommendation algorithm provides an indication whether a pair of granularities can be meaningfully examined together (a "harmony"), or when they cannot (a "clash").

 Time granularities can be used to create data visualizations to explore for periodicities, associations and anomalies. The granularities form categorical variables (ordered or unordered) which induce groupings of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables. A recommendation of appropriate distribution displays is provided.

 The methods are implemented in the open source R package `gravitas`, providing functions for creating granularities and exploring the associated time series which are consistent with a tidy workflow (@Grolemund2018-po), and the probability distributions can be examined using the range of graphics available in `ggplot2` [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \usepackage{mathtools,amssymb,booktabs,amsthm}
  \def\mod{~\text{mod}~}
  \newtheorem{definition}{Definition}
  \newtheorem*{example}{Example}
  \usepackage{mathptmx}
  \usepackage{caption}
  \DeclareCaptionStyle{italic}{labelfont={bf},textfont={it},labelsep=colon}
  \captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
  \captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "figure/",
  fig.align = "center",
  fig.show = "hold",
  cache = TRUE,
  cache.path = "cache/",
  out.width = ifelse(is_html_output(), "100%", "\\textwidth")
)
```

@@

```{r load}
```

@@

@@

@@

@@

This work provides tools for systematically exploring bivariate granularities within the tidy workflow. In particular, we

@@

@@

@@

\begin{definition}\label{def:definition}
A \textbf{time domain} is a pair $(T; \le)$ where $T$ is a non-empty set of time instants and $\le$ is a total order on $T$.
\end{definition}

\noindent The time domain is assumed to be *discrete*, and there is unique predecessor and successor for every element in the time domain except for the first and last.

\begin{definition}\label{def:index set}
There is a unique \textbf{index set}, $Z=\{z: z \in \mathbb{Z}_{\geq 0}\}$, which map the time instants to a set of non-negative integers. Often, this is thought of as $t=0, \dots, T$.
\end{definition}

\begin{definition}\label{def:linear}
A \textbf{linear granularity} is a mapping $G$ from the integers (the index set, $Z$) to subsets of the time domain such that:
  (1) if $i < j$ and $G(i)$ and $G(j)$ are non-empty, then each element of $G(i)$ is less than all elements of $G(j)$; and
  (2) if $i < k < j$ and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty.
Each non-empty subset $G(i)$ is called a \textbf{granule}.
\end{definition}

\noindent This implies that the granules in a linear granularity are non-overlapping, continuous and ordering is maintained. The indexing for each granule can also be associated with textual representation, called the label. A discrete time model often uses a fixed smallest linear granularity named by @Bettini1998-ed **bottom granularity**. \autoref{fig:linear-time} illustrates the linear time granularities. Here, "hour" is the bottom granularity and "day", "week", "month" and "year" are linear granularities which maps from index set to subsets of the hourly time domain. If we have "hour" running from $\{0, 1, \dots,t\}$, we will have "day" running from $\{0, 1, \dots, \lfloor t/24\rfloor\}$. These linear granularities are ordered with ordering guided by the index set which is a set of integers. Hence, they are uni-directional and non-repeating.

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="(ref:linear-time)", fig.pos="!htb"}
```

@@

@@

\begin{definition}\label{def:finerthan}
A linear granularity $G$ is \textbf{finer than} a linear granularity $H$, denoted $G \preceq H$, if for each index $i$, there exists an index $j$ such that
$G(i) \subset H(j).$
\end{definition}

\begin{definition}\label{def:groupsinto}
A linear granularity $G$ \textbf{groups into} a linear granularity $H$, denoted
$G \trianglelefteq H$, if for each index $j$ there exists a (possibly infinite) subset $S$ of the integers such that $H(j) = \bigcup_{i \in S}G(i).$
\end{definition}

\noindent For example, both $day \trianglelefteq week$ and $day \preceq week$ hold, since every granule of $week$ is the union of some set of granules of day and each day is a subset of a $week$. Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year, if leap years are ignored. This grouping period becomes 4 and 400 years with the inclusion of leap years and leap seconds respectively.

\begin{definition}\label{def:periodical}
A granularity $G$ is \textbf{periodical} with respect to a granularity $H$ if:
(1) $G \trianglelefteq H$; and
(2) there exist $R$, $P \in \mathbb{Z}_+$, where $R$ is less than the number of granules of $H$, such that for all $i \in \mathbb{Z}$, if $H(i) = \bigcup_{j \in S}G(j)$ and $H (i + R) \neq \phi$ then $H (i + R) = \bigcup_{j \in S} G(j + P)$.
\end{definition}

@@

For example, in a Gregorian calendar without leap years we could say $day$ groups periodically into month with the period $P = 365$ and the number of granules of $month$ in each period given by $R=12$. In a Gregorian calendar with leap years, $day$ groups quasi-periodically into $month$ with the exceptions of the time domain corresponding to $29^{\text{th}}$ February of any year.

\begin{definition}\label{def:order}
The \textbf{order} of a linear granularity is the level of coarseness associated with a linear granularity. A linear granularity G will have lower order than H if each granule of G is composed of lower number of granules of bottom granularity than each granule of H.
\end{definition}

With two linear granularities $G$ and $H$, if $G$ *groups into* or *finer than* $H$ then $G$ is of lower order than $H$. For example, if the bottom granularity is day, then granularity week will have lower order than month since each week consist of fewer days than each month. Here neither week *groups into* month or week *finer than* month, but their relative order can be determined.

@@

\begin{definition}\label{def:grouping_operation}
Let $G_1$ be a full-integer labelled granularity, and $m$ a positive integer. \textbf{The grouping operation} Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labelled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$
The grouping operation Group$_m(G)$, provides a new granularity whose granules are the integer part of the quotient when the integer set spanned by linear granularity $G$ is divided by $m$, such that, $G_2(i) = \lfloor G_1(j)/m \rfloor$.
\end{definition}

@@

@@

@@

Cyclic granularities represent cyclical repetitions in time. They can be thought of as additional categorizations of time that are not linear. Cyclic granularities can be constructed by operating on two linear granularities. Cycles can be either _regular_, called **circular**, or _irregular_, **quasi-circular** when these two linear granularities relate periodically.

@@

\begin{definition}\label{def:circular}
A \textbf{circular granularity} $C_{B, G}$ relates a linear granularity $G$ to the bottom granularity B, if
\begin{equation} \label{eq:circular-gran}
\begin{split}
C_{B, G}(z) & = z\mod P(B, G) \quad \forall z \in \mathbb{Z}_{\geq 0} \\
\end{split}
\end{equation}
where
z denotes the index set,
B denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G$ with regular mapping, and $P \equiv P(B, G)$ is the number of granules of $B$ in each granule of $G$, also called the period of the grouping $(B, G)$.
\end{definition}

For example, \autoref{fig:circular-dow} illustrates the linear and the corresponding cyclical granularities. Cyclical granularities can be considered to be cutting the linear granularity into pieces, and stacking them to match the cycles (as shown in b). $B, G, H$ (day, week, fortnight, respectively) are linear granularities. The circular granularity $C_{B , G}$ (day-of-week)is constructed from $B$ and $G$. The circular granularity $C_{B, H}$ (day-of-fortnight) is constructed from $B$ and $H$. These overlapping cyclical granularities share elements from the linear granularity. Each of $C_{B , G}$ and $C_{B , H}$ consist of repeated patterns $\{0, 1, \dots, 6\}$ and $\{0, 1, \dots, 13\}$ with $P=7$ and $P=14$ respectively. Each circular granularity can use descriptive label mappings. Suppose ${L}$ is a label mapping that defines an unique label for each index $\ell \in \{ 0,1,\dots, (P-1)\}$, then the label mapping $L$ for $C_{B, G}$ can be defined (for example) as
\begin{align*}
  L: \{0,1, \dots, 6\} & \longmapsto\ \{\text{Sun}, \text{Mon}, \dots, \text{Sat}\}\\
\text{or}\qquad
  L: \{0,1, \dots, 6\} & \longmapsto\ \{\text{Sunday}, \text{Monday}, \dots, \text{Saturday}\}.
\end{align*}

```{r circular-dow,echo=FALSE,out.width = "100%",fig.cap= "(ref:circular-dow)",fig.pos="!htb"}
```

@@

In general, any circular granularity relating two linear granularity can be expressed as $C_{(G, H)}(z) = \lfloor z/P(B,G) \rfloor\mod P(G,H)$, where linear granularity $H$ is periodic with respect to linear granularity $G$ with regular mapping and $P(G,H)$ is the period of the grouping $(G, H)$. Table \ref{tab:definitions} shows representation of circular granularities $C_i$ relating two linear granularities with period $P_i$ and minutes as the bottom granularity.
